---
title: "BIC MIC Model"
author: "Raynah Cheng"
date: "2024-12-03"
output: pdf_document
---
```{r, include=FALSE}
source <- 
  read.csv("kc61_sample_result_age_source.csv")

dim(source)

#columns of interest: Indicator (age), Borderline changes, mild, moderate,
#and severe dyskaryosis

#only column that has NA's is severe dyskaryosis; sub in values for NA to clean
source$Severe_dyskaryosis <- gsub("not included", NA, source$Severe_dyskaryosis)
sum(is.na(source$Severe_dyskaryosis) == TRUE)

uniq_ind <- unique(source$Indicator)
for (i in 1:length(uniq_ind)){
  source$Indicator <- gsub(uniq_ind[i], i, source$Indicator)
} 
source$Indicator <- as.numeric(source$Indicator)
#changing each age group to correspond with a number from 1-19. for example, <20 is 1

#removing NAs and converting to numeric values
severe_NA <- source[is.na(source$Severe_dyskaryosis) == FALSE, ]
severe_NA$Severe_dyskaryosis <- as.numeric(severe_NA$Severe_dyskaryosis)

comparison_data <- source[,c("Borderline_changes", "Mild_dyskaryosis", "Moderate_dyskaryosis","Severe_dyskaryosis")]
comparison_data$Severe_dyskaryosis <- as.numeric(comparison_data$Severe_dyskaryosis)
```
**BIC MIC Models** 
Raynah Cheng

-cleaning data taken from Project Check in 2

```{r}
#splitting data into test and train
train_index <- 1:165
test_index <- 166:nrow(comparison_data)

train_frame <- comparison_data[train_index,]
test_frame <- comparison_data[test_index,]


train_model <- lm(Borderline_changes ~ ., data = train_frame)
summary(train_model)

back_BIC <- step(train_model, direction = "backward", 
                 k = log(nrow(train_frame)), trace = 0)
summary(back_BIC)

back_MIC <- step(train_model, direction = "backward", 
                 k = nrow(train_frame), trace = 0)
summary(back_MIC)
```
The BIC model chooses all of the dyskaryosis variables, with equal about equal levels of significance. However, we can see that the severe dyskaryosis t value is higher, at 8.748, compared to the others at around 4.5. For our BIC model, 81.12% of the variance can be explained by our model. The MIC model only chooses the severe dyskaryosis variable, at a .001 level of signifiance, with a much higher t value than the BIC model. However, the variance that can be explained goes down a little bit, at 78.69%. 

```{r}
test_MIC <- predict(back_MIC, test_frame)
test_BIC <- predict(back_BIC, test_frame)

cor(test_MIC, test_frame$Borderline_changes, use = "complete.obs")^2 #r-squared for back_MIC model
cor(test_BIC, test_frame$Borderline_changes, use = "complete.obs")^2 #r-squared for back_BIC model

errors_MIC <- test_frame$Borderline_changes - test_MIC
errors_BIC <- test_frame$Borderline_changes- test_BIC

sqrt(mean(errors_MIC^2, na.rm = TRUE)) #RMSE for back_MIC
sqrt(mean(errors_BIC^2, na.rm = TRUE)) #RMSE for back_BIC

mean(abs(errors_MIC), na.rm = TRUE) #MAE for back_MIC
mean(abs(errors_BIC), na.rm = TRUE) #MAE for back_BIC
```
| Model  | $R^2$ | RMSE | MAE | 
|--------|-------|-------|--------|
|   MIC  |  .997  |  1763.82   |  1286.17  |
|   BIC  |  .991   |  1650.03  |  1056.62  |

The BIC model has a lower R^2 value, but the errors are lower. The MIC model has a higher R^2 value, but the errors are higher. I think that we would pick the BIC model because the R^2 is not that much lower and it predicts our data a lot better.